Finiteness in Cubical Type Theory

### 2020-01-23 paper.tex
more

\section{Countably Infinite Types}
In the previous sections we saw different flavours of finiteness which were
really just different flavours of relations to \(\mathbf{Fin}\).
In this section we will see that we can construct a similar classification of
relations to \(\mathbb{N}\), in the form of the countably infinite types.
\subsection{Split Countable Types}
Our first foray into the world of countable types will be a straightforward
analogue to the split enumerable types.
We need change only one element: instead of a support \emph{list}, we instead
have a support \emph{stream}, which is its infinite.
\begin{romdefinition}[Stream]
  \begin{equation}
    \mathbf{Stream}(A) = \llbracket \top \triangleright \text{const} (\mathbb{N}) \rrbracket(A) \simeq \mathbb{N} \rightarrow A
  \end{equation}
\end{romdefinition}
\begin{romdefinition}[Split Countability]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \mathbf{Stream}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
\end{romdefinition}
This type is definitionally equal to it surjection equivalent (\(\mathbb{N}
\twoheadrightarrow ! \; A\)).
\subsubsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
\input{figures/pairing-functions.tex}
\begin{romtheorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
\end{romtheorem}
\begin{proof}
  The main task here is to figure out a pattern which will pair elements of the
  two support streams without diverging.
  Figure~\ref{cartesian} illustrates why the pairing function we used for lists
  won't work: it needs to exhaust one of the input lists before inspecting
  anything other than the first element of the other.
  Of course, we can't exhaust a stream, so this diverges.

  Instead, we will use the pattern in figure~\ref{cantor}.
  We will produce a stream of lists (where each list is a diagonal), which is
  then concatenated to the final output stream.
  The intermediate stream is sometimes called the discrete convolution of the
  two input streams.
\end{proof}
\subsubsection{Kleene Star}
While we lose some closures with the inclusion of infinite types, we gain some
others.
In particular, we have the Kleene star.
This means that we have closure under lists.
\begin{romtheorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{romtheorem}
\begin{proof}
  As with the proof of closure under \(\Sigma\), our main task here is to figure
  out a pattern that reaches any given finite list in finite time.
  Our first step is to split up the problem as we did for \(\Sigma\)-closure: we
  produce an intermediate stream of lists, which we then concatenate to give the
  final result.
  The intermediate stream is constructed like so:
  \begin{equation}
    \mathit{xs}^\star_i = \left\{ \text{map}(\mathit{xs} , x) \mid \mathit{x} \in \mathbf{List}(\mathbb{N}) , \text{sum}(\text{map}(1+, x)) \equiv i \right\}
  \end{equation}
  In other words, the \(i\)th element is all lists such that the sum of the
  successor of the indices in the list is equal to \(i\).
\end{proof}
\subsection{Other Types of Countability}
There are of course other forms of countability available to us: manifest
countability, cardinal, etc.
We have a similar structure of relationships to the finite types, with the
exception that we cannot construct an isomorphism with \(\mathbb{N}\) fiven a
surjection from \(\mathbb{N}\).
\begin{itemize}
  \item Definition
  \item \(\Sigma\)
  \item Kleene star
  \item Truncated
  \item Topos
\end{itemize}
\section{Search}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{romdefinition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{romdefinition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.

>>>>

We have proven the above statements for both \(\mathit{Set}\) and
\(\mathit{FinSet}\).
As far as we know, this is the first formalisation of either.
\section{Countably Infinite Types}
In the previous sections we saw different flavours of finiteness which were
really just different flavours of relations to \(\mathbf{Fin}\).
In this section we will see that we can construct a similar classification of
relations to \(\mathbb{N}\), in the form of the countably infinite types.
\subsection{Two Countable Types}
The two types for countability we will consider are analogous to split
enumerability and cardinal finiteness.
The change will be a simple one: in terms of surjections and bijections, we will
swap out \(\mathbf{Fin}\) for \(\mathbb{N}\).
In terms of support lists and cover proofs, we will swap out lists for streams.
A stream can be defined many ways in type theory: we take rather a low-tech
approach here, saying that a stream is a function from the natural numbers.
\begin{romdefinition}[Stream]
  \begin{equation}
    \mathbf{Stream}(A) \coloneqq \mathbb{N} \rightarrow A
  \end{equation}
  Streams are also definable as containers:
  \begin{equation}
    \mathbf{Stream}(A) \simeq \llbracket \top , \text{const}(\mathbb{N}) \rrbracket
  \end{equation}
  However the two definitions are clearly equivalent, and furthermore it is
  simpler to work with the first.
\end{romdefinition}

From this, we can define ``split countability''.
\begin{romdefinition}[Split Countability]
  \begin{equation}
    \mathcal{E}!(A) \coloneqq \Sigma {(\mathit{xs} : \mathbf{Stream}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
\end{romdefinition}
This type is definitionally equal to it surjection equivalent (\(\mathbb{N}
\twoheadrightarrow ! \; A\)).
We construct the unordered, propositional version of the predicate in much the
same way as we constructed cardinal finiteness.
\begin{romdefinition}[Countability]
  \begin{equation}
    \mathcal{E}(A) \coloneqq \lVert \mathcal{E}!(A) \rVert
  \end{equation}
\end{romdefinition}

From both of these types we can derive decidable equality.
\begin{romlemma}
  Any countable type has decidable equality.
\end{romlemma}
\subsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
\begin{romtheorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
\end{romtheorem}
\begin{romtheorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{romtheorem}
\subsection{Categories}
\section{Search}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{romdefinition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{romdefinition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.
